---
title: 'MIS 64060: Assignment_3: Naive Bayes for classification '
author: "Eyob Tadele"
date: "10/15/2021"
output:
  pdf_document: default
  html_document: default
---

### Project Objective

The purpose of this assignment is to use Naive Bayes for classification.

The file UniversalBank.csv contains data on 5000 customers of Universal Bank. The data include
customer demographic information (age, income, etc.), the customer’s relationship with the bank
(mortgage, securities account, etc.), and the customer response to the last personal loan campaign
(Personal Loan). Among these 5000 customers, only 480 (= 9.6%) accepted the personal loan that
was offered to them in the earlier campaign. In this exercise, we focus on two predictors: Online
(whether or not the customer is an active user of online banking services) and Credit Card
(abbreviated CC below) (does the customer hold a credit card issued by the bank), and the outcome
Personal Loan (abbreviated Loan below).

### Importing a UniversalBank.csv dataset into r, load relevant libraries, and printout stats about the data.

```{r}
library(caret)
library(e1071)

univBank <- read.csv("UniversalBank.csv")
str(univBank)
head(univBank)
```
### Transforming the two predictors(Online,Credit Card) and the outcome(Personal.Loan) into factors.
```{r}
univBank$CreditCard <- factor(univBank$CreditCard)
univBank$Online <- factor(univBank$Online)
univBank$Personal.Loan <- factor(univBank$Personal.Loan)
str(univBank)
```
### Partition the data into training (60%) and validation (40%) sets.
```{r}
selVars <- c(10,13,14)    # selecting Online, Credit Card, and Personal.Loan 
set.seed(420)
univBank.part.mod <- createDataPartition(univBank$Personal.Loan,p=0.6, list = FALSE)
univBank.tr <- univBank[univBank.part.mod,selVars]
univBank.va <- univBank[-univBank.part.mod,selVars]
summary(univBank.tr)
summary(univBank.va)

```

### Part A: Create a pivot table for the training data with Online as a column variable, CC as a row variable, and Loan as a secondary row variable.
```{r}
# creating a pibot table using the ftable() function
attach(univBank.tr)
ftable(Personal.Loan,CreditCard,Online)
detach(univBank.tr)
```

### Part B: Looking at the pivot table, calculate the probability of loan offer acceptance for a customer who owns a bank credit card and is an active online banking services user. 
This is the probability of loan acceptance (Loan = 1) conditional on
having a bank credit card (CC = 1) and being an active user of online banking services (Online = 1). 

P(L=1 | C=1,O=1) = 52/(52+476) = 0.0984

### Part C: Create two separate pivot tables for the training data. One will have Loan (rows) as a function of Online (columns) and the other will have Loan (rows) as a function of CC.

```{r}
attach(univBank.tr)
ftable(Online,Personal.Loan)
ftable(CreditCard,Personal.Loan)
detach(univBank.tr)
```

### Part D: Compute the conditional probabilities of: 

### i. P(CC = 1 | Loan = 1) (the proportion of credit card holders among the loan acceptors)
    P(CC = 1 | Loan = 1) = 91/288 =  0.3159
  
### ii. P(Online = 1 | Loan = 1)
    P(Online = 1 | Loan = 1) = 177/288 = 0.6145
  
### iii. P(Loan = 1) (the proportion of loan acceptors)
  P(Loan = 1) = 288/3000 = 0.096
  
### iv. P(CC = 1 | Loan = 0)
  P(CC = 1 | Loan = 0) = 799/2712 = 0.2946

### v. P(Online = 1 | Loan = 0)
  P(Online = 1 | Loan = 0) = 1606/2712 = 0.5921

### vi. P(Loan = 0)
  P(Loan = 0) = 2712/3000 = 0.904
  
### Part E: Use the quantities computed above to compute the naive Bayes probability P(Loan = 1 | CC = 1, Online = 1).
P(L=1|CC=1,O=1) = P(CC=1|L=1) * P(O=1|L=1) * P(L=1) / 
                  P(CC=1|L=1) * P(O=1|L=1) * P(L=1) + P(CC=1|L=0) * P(O=1|L=0) * P(L=0) 
                = (91/288)(177/288)(288/3000) / ((91/288)(177/288)(288/3000) + (799/2712)(1606/2712)(2712/3000))

P(L=1|CC=1,O=1) = 0.1057

### Part F: Compare this value with the one obtained from the pivot table in (B). Which is a more accurate estimate?
The value obtained through naive Bayes method is 0.1057, and 0.0984 using direct method. Naive Bayes makes 
an assumption that all the features are independent of eachother and that may not be the case in real lifecycle. 
Making that assumption may make that less accurate.

### Part G: Which of the entries in this table are needed for computing P(Loan = 1 | CC = 1, Online = 1)?
The entries needed are: P(CC=1|L=1), P(O=1|L=1), P(L=1), P(CC=1|L=0), P(O=1|L=0), P(L=0)

### Part G: Run naive Bayes on the data. Examine the model output on training data, and find the entry that corresponds to P(Loan = 1 | CC = 1, Online = 1). Compare this to the number you obtained in (E). 

```{r}
univBank.nb <- naiveBayes(Personal.Loan ~ ., data = univBank.tr)  # using the naiveBayes model
univBank.nb
```

**Calculating values from the naiveBayes model probability tables results in a value of 0.1057 : **
  **(0.3159722)(0.6145833)(0.096) / ((0.3159722)(0.6145833)(0.096)+(0.2946165)(0.5921829)(0.904)) = 0.1057**

**This is the same as the value obtained in E.** 
  
  